Adaptive global optimization using graphics accelerators\thanks{This study was supported by the Russian Science Foundation, project No.\,16-11-10150.}

### 2020-04-12 2020/RSD_GPU/RSD2020.tex
Первая версия, переведенного текста в tex.

\documentclass{svproc}
%
% RECOMMENDED 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%
%
\usepackage{graphicx}
\usepackage{marvosym}

@@

\maketitle              % typeset the title of the contribution

@@

A promising direction in the field of parallel global optimization (which, indeed, is true in many areas related to the software implementation of time-consuming algorithms) is the use of graphics processing units (GPUs). In the past decade, graphics accelerators have rapidly increased performance to meet the ever-growing demands of graphics application developers. Additionally, in the past few years some principles for developing graphics hardware have changed, and as a result it has become more programmable. Today, a graphics accelerator is a flexibly programmable, massive parallel processor with high performance, which is in demand for solving a range of computationally time-consuming problems \cite{Hwu2011}.

However, the potential for graphics accelerators to solve global optimization problems has not yet been fully realized. Using the GPUs, they basically parallelize nature-inspired optimization algorithms, which are somehow based on the idea of random search (see, for example, \cite{Ferreiro2013,Garcia2014,Langdon2011}). By virtue of their stochastic nature, algorithms of this type guarantee convergence to the global minimum only in the sense of probability, which differentiates them unfavorably from deterministic methods.

With regard to many deterministic algorithms of Lipschitzian global optimization with guaranteed convergence, parallel variants have been proposed \cite{Evtushenko2009,He2008,Paulavicius2011}. However, these versions of algorithms are parallelized on the CPU using shared and/or distributed memory; presently, no GPU implementations have been made. For example, \cite{Paulavicius2011} describes parallelization of an algorithm based on the ideas of the branch and boundary method using MPI and OpenMP.

Within the framework of this research, we consider the problems of capturing the optimum, which are characterized by a lengthy period for calculating the values of objective function in comparison with the time needed for processing them. For example, objective function can be specified using systems of linear algebraic equations, systems of ordinary differential equations, etc. Currently, graphics accelerators can be used to solve problems of this type. Moreover, an accelerator can solve several such problems at once \cite{Kindratenko2014}; i.e., using the GPU, one can calculate multiple function values simultaneously.

@@

\section{Multidimensional parallel global search algorithm}

\section{Parallel global search algorithm}

@@

@@

@@

@@

@@

@@

%
% ---- Bibliography ----
%
\bibliographystyle{spmpsci}
\bibliography{bibliography}{}

>>>>

\documentclass{svproc}
%
% RECOMMENDED 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%
%
\usepackage{graphicx}
\usepackage{marvosym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[russian]{babel}

@@

%\maketitle              % typeset the title of the contribution

@@

A promising direction in the field of parallel global optimization (which, indeed, is true in many areas related to the software implementation of time-consuming algorithms) is the use of graphics processing units (GPUs). In the past decade, graphics accelerators have rapidly increased performance to meet the ever-growing demands of graphics application developers. Additionally, in the past few years some principles for developing graphics hardware have changed, and as a result it has become more programmable. Today, a graphics accelerator is a flexibly programmable, massive parallel processor with high performance, which is in demand for solving a range of computationally time-consuming problems \cite{Ref1}.

However, the potential for graphics accelerators to solve global optimization problems has not yet been fully realized. Using the GPUs, they basically parallelize nature-inspired optimization algorithms, which are somehow based on the idea of random search (see, for example, \cite{Ref2,Ref3,Ref4}). By virtue of their stochastic nature, algorithms of this type guarantee convergence to the global minimum only in the sense of probability, which differentiates them unfavorably from deterministic methods.

With regard to many deterministic algorithms of Lipschitzian global optimization with guaranteed convergence, parallel variants have been proposed \cite{Ref5,Ref6,Ref7}. However, these versions of algorithms are parallelized on the CPU using shared and/or distributed memory; presently, no GPU implementations have been made. For example, \cite{Ref7} describes parallelization of an algorithm based on the ideas of the branch and boundary method using MPI and OpenMP.

Within the framework of this research, we consider the problems of capturing the optimum, which are characterized by a lengthy period for calculating the values of objective function in comparison with the time needed for processing them. For example, objective function can be specified using systems of linear algebraic equations, systems of ordinary differential equations, etc. Currently, graphics accelerators can be used to solve problems of this type. Moreover, an accelerator can solve several such problems at once \cite{Ref8}; i.e., using the GPU, one can calculate multiple function values simultaneously.

@@

\section{Multidimensional parallel global search algorithm}
Consider the problem of finding the global minimum of the \(N\)-dimensional function \(\phi(y)\) in the hyperinterval  \(D=\{y\in R^N:a_i\leqslant x_i\leqslant{b_i}, 1\leqslant{i}\leqslant{N}\}\)  We will assume that the function satisfies the Lipschitz condition with an a priori unknown constant \(L\).
\begin{equation}
\label{task}
\phi(y^*)=\min\{\phi(y):y\in D\}
\end{equation}
\begin{equation}
\label{lip}
|\phi(y_1)-\phi(y_2)|\leqslant L\Vert y_1-y_2\Vert,y_1,y_2\in D,0<L<\infty
\end{equation}

In this paper, we will use an approach based on the idea of reducing dimensionality using the Peano space-filling curve \(y(x)\), which continuously and unambiguously maps a segment of the real axis \([0,1]\) onto an \(n\)-dimensional cube
\begin{equation}
\label{cube}
\lbrace y\in R^N:-2^{-1}\leqslant y_i\leqslant 2^{-1},1\leqslant i\leqslant N\rbrace=\{y(x):0\leqslant x\leqslant 1\}
\end{equation}

The questions of numerical construction of approximations to Piano curve (evolvents) and the corresponding theory are discussed in detail in \cite{Ref9}, \cite{Ref10}. Using evolvents \(y(x)\) reduces the multidimensional problem (\ref{task}) to a one-dimensional problem
\begin{displaymath}
\label{oneDimTask}
\phi(y^*)=\phi(y(x^*))=\min\{\phi(y(x)):x\in [0,1]\}
\end{displaymath}

An important property is that the relative differences of the function remain limited: if the function \(\phi(y)\) in the region \(D
\) satisfies the Lipschitz condition, then the function \(\phi(y(x))\) in the interval \([0,1]\) will satisfy the uniform Holder condition
\begin{displaymath}
\label{holder}
|\phi(y(x_1))-\phi(y(x_2))|\leqslant H{|x_1-x_2|}^{\frac{1}{N}},x_1,x_2\in[0,1]
\end{displaymath}
where the Holder constant \(H\) is related to the Lipschitz constant \(L\) by the ratio
\begin{displaymath}
H=4Ld\sqrt{N},d=\max\{b_i-a_i:1\leqslant i\leqslant N\}
\end{displaymath}

Therefore, without limiting generality, we can consider minimizing the one-dimensional function \(f(x)=\phi(y(x)), x\in[0,1]\), which satisfies the Holder condition.

The considered algorithm for solving this problem involves constructing a sequence of points \(x_k\), in which the values of objective function \(z_k = f(x_k)\) are calculated. We will call the process of computing the value of a function at a single point a trial. Assume that we have \(P\geqslant 1\) computational elements at our disposal and \(P\) tests which are performed simultaneously (synchronously) within a single iteration of the method. Let \(k(n)\) denote the total number of tests performed, after \(n\) parallel iterations.

In the first iteration of the method, the test is carried out at an arbitrary internal point \(x^1\) of the interval \([0,1]\). Let \(n>1\) iterations of the method be performed, during which tests were carried out at \(k = k(n)\) points \(x^i, 1\leqslant i\leqslant k\). Then the points  \(x^{k+1},\dotsc,x^{k+p}\). search tests of the following \((n+1)\)-th iteration are determined in accordance with the rules:

Step 1. Renumber the points of the set \(X_k=\{x^1,\dotsc,x^k\}\cup\{0\}\cup\{1\}\), which includes the boundary points of the interval \([0,1]\), as well as the points of the previous tests, with the lower indices in the order of their increasing coordinate values, i.e.
\begin{displaymath}
0=x_0<x_1<\dotsc<x_{k+1}=1
\end{displaymath}

Step 2. Assuming \(z_i=f(y(x_i)),1\leqslant i\leqslant k\) calculate the values
\begin{displaymath}
\label{step2}
\mu=\max_{1\leqslant i\leqslant k}\dfrac{|z_i-z_{i-1}|}{\Delta_i},
\begin{matrix}
    M = 
    \left\{
    \begin{matrix}
    r\mu,\mu>0 \\
    1,\mu=0
    \end{matrix} \right.
    \end{matrix}
\end{displaymath}
where \(r > 1\) is the specified parameter of the method, and \(\Delta_i=(x_i-x_{i-1})^\frac{1}{N}\).

Step 3. For each interval \((x_{i-1},x_i),1\leqslant i\leqslant k+1\), calculate the characteristic in accordance with the formulas 
\begin{displaymath}
\label{step3_1}
R(1)=2\Delta_1-4\dfrac{z_1}{M},R(k+1)=2\Delta_{k+1}-4\dfrac{z_k}{M}
\end{displaymath}
\begin{displaymath}
\label{step3_2}
R(i)=\Delta_i+\dfrac{(z_i-z_{i-1})^2}{M^2\Delta_i}-2\dfrac{z_i+z_{i-1}}{M},1<i<k+1
\end{displaymath}

Step 4. The characteristics \(R(i),1\leqslant i\leqslant k+1\), arrange in descending order 
\begin{displaymath}
\label{step4}
R(t_1)\geqslant R(t_2)\geqslant \dots \geqslant R(t_{k})\geqslant R(t_{k+1})
\end{displaymath}
and select \(P\) of the largest characteristics with interval numbers \(t_j,1\leqslant j\leqslant P\).

Step 5. Carry out new tests at the points \(x_{k+j},1\leqslant j\leqslant P\), calculated using the formulas
\begin{displaymath}
x_{k+j}=\dfrac{x_{t_j}+x_{t_j-1}}{2},t_j=1,t_j=k+1
\end{displaymath}
\begin{displaymath}
\label{step5}
{{x}^{k+1}}=\frac{{{x}_{{{t}_{j}}}}+{{x}_{{{t}_{j}}-1}}}{2}-\text{sign}({{z}_{{{t}_{j}}}}-{{z}_{{{t}_{j}}-1}})\frac{1}{2r}{{\left[ \frac{|{{z}_{{{t}_{j}}}}-{{z}_{{{t}_{j}}-1}}|}{\mu } \right]}^{N}}, 1<{{t}_{j}}<k+1
\end{displaymath}

The algorithm stops working if the condition \(\Delta_{t_j}\leqslant \varepsilon\)  is satisfied for at least one number \(t_j,1\leqslant j\leqslant P\); here \(\varepsilon>0\)  is the specified accuracy. As an estimate of the globally optimal solution to the problem (\ref{task}), the values are selected 
\begin{displaymath}
f_{k}^{*}=\underset{1\le i\le k}{\mathop{\min }}\,f({{x}^{i}}),   x_{k}^{*}=\arg \quad \underset{1\le i\le k}{\mathop{\min }}\,f({{x}^{i}})
\end{displaymath}

For the rationale in using this method of organizing parallel computing, see \cite{Ref11}. Modifications that account for the presence of inequality constraints in the problem, as well as information about the derivative of the objective function, are presented in \cite{Ref12} – \cite{Ref14}.

@@

Reducing multidimensional problems to one-dimensional ones through the use of evolvents has important properties such as continuity and preserving uniform bounding of function differences with limited argument variation. However, some information about the proximity of points in multidimensional space is lost, since the point \(x\in [0,1]\) has only left and right neighbors, and the corresponding point \(y(x) \in R^N\) has neighbors in \(2^N\) directions. As a result, when using evolvents, the images \(y' , y''\) that are close in \(N\)-dimensional space can correspond to rather distant prototypes \(x' , x''\) on the interval \([0,1]\). This property leads to redundant calculations, because several limit points \(x' , x''\) of the test sequence generated by the method on the segment \([0,1]\), can correspond to a single limit point \(y\) in \(N\)-dimensional space.

A possible way to overcome this disadvantage is to use a set of evolvents (multiple mappings)

\begin{displaymath}
\label{eq:142}
Y_L(x)=\left\{y^0(x),\ y^1(x),...,\ y^L(x)\right\}
\end{displaymath}

instead of using a single Peano curve \(y(x)\) (see \cite{Ref19}, \cite{Ref20}). For example, each Peano curve \(y^i(x)\) from \(Y_L(x)\) can be obtained as a result of some shift along the main diagonal of the hyperinterval \(D\), the original curve \(y(x)\). Another way is to rotate the sweep around the coordinate. The set of evolvents that have been constructed allows us to obtain for any close images \(y', y''\), close prototypes \(x', x''\) for some mapping \(y^i(x)\).

   Using a set of mappings leads to the formation of a corresponding set of one-dimensional multiextremal problems

\begin{displaymath}
\label{6_problem_l}
\min{\left\{\varphi(y^l(x)):x\in [0,1], \;  \right\}}, \ 0 \leq l \leq L.
\end{displaymath}

Each problem from this set can be solved independently, and any calculated value \(z= \varphi(y')\), \(y'=y^i(x')\) of the function \(\varphi(y)\) in the \(i\)-th problem can be interpreted as calculating the value \(z= \varphi(y')\), \(y'=y^s(x'')\) for any other \(s\)-th problem without repeated labor-intensive calculations of the function \(\varphi(y)\). Such informational unity makes it possible to solve the entire set of problems in a parallel fashion. This approach was discussed in detail in \cite{Ref21}.

\subsection{Recursive dimensionality reduction scheme}

The recursive optimization scheme is based on the well-known relation \cite{Ref22}

\begin{equation}
\label{nestedScheme}
\min{\phi(y):y\in D}=\min_{a_1\leqslant y_1\leqslant b_1}\min_{a_2\leqslant y_2\leqslant b_2}\dots\min_{a_1\leqslant y_N\leqslant b_N}\phi(y)
\end{equation}

which allows one to replace the solution of the multidimensional problem (\ref{task}) with the solution of a family of one-dimensional subproblems recursively related to each other.
Let's introduce a set of functions

\begin{equation}
\label{phiN}
\phi_N(y_1,\dots,y_N)=\phi(y_1,\dots,y_N)
\end{equation}
\begin{equation}
\label{phiI_}
\phi_i(y_1,\dots,y_i)=\min_{a_{i+1}\leqslant y_{i+1} \leqslant b_{i+1}}\phi_{i+1}(y_1,\dots,y_i,y_{i+1}),1\leqslant i\leqslant N-1
\end{equation}

Then, in accordance with the relation (\ref{nestedScheme}) , the solution of the original problem (\ref{task}) is reduced to the solution of a one-dimensional problem

\begin{equation}
\label{phiFirst}
\phi_1(y_1^*)=\min\{\phi_1(y_1),y_1\in [a,b]\}
\end{equation}

However, each calculation of the value of the one-dimensional function  \(\phi_1(y_1)\) at some fixed point corresponds to the solution of a one-dimensional minimization problem

\begin{displaymath}
\label{phi2}
\phi_2(y_1,y_2^*)=\min\{\phi(y_1,y_2):y_2\in [a_2,b_2]\}
\end{displaymath}

And so on, until the calculation   according to (\ref{phiN}).

For the recursive scheme described above, a generalization (block recursive scheme) is proposed that combines the use of sweeps and a recursive scheme in order to efficiently parallelize computations.

Consider the vector y as a vector of block variables
\begin{displaymath}
y=(y_1,\dots,y_N)=(u_1,u_2,\dots,u_M)
\end{displaymath}

Where the \(i\)-th block variable \(u_i\) is a dimensional vector   of sequentially taken components of the vector \(y\), i.e. \(u_1=(y_1,y_2,\dots,y_{N_1}),u_2=(y_{N_1+1},y_{N_1+2},\dots,y_{N_1+N_2}),\dots,u_M=(y_{N-N_M+1},y_{N-N_M+2},\dots,y_N)\), while \(N_1+N_2+\dots+N_M=N\).

Using new variables, the main relation of the multistep scheme (\ref{nestedScheme}) can be rewritten as

\begin{equation}
\label{blockNested}
\min\phi(y)_{y\in D}=\min_{u_1\in D_1}\min_{u_2\in D_2}\dots\min_{u_M\in D_M}\phi(y)
\end{equation}

where the subdomains  \(D_i,1\leqslant i\leqslant M\), are projections of the original search domain \(D\) onto the subspaces corresponding to the variables \(u_1,1\leqslant i\leqslant M\) .

The formulas that determine the method for solving problem (\ref{task}) based on relations (\ref{blockNested}) generally coincide with the recursive scheme (\ref{phiN})-(\ref{phiFirst}). One need only replace the original variables  \(y_i,1\leqslant i\leqslant N\), with block variables  \(u_1,1\leqslant i\leqslant M\). 
In this case, the fundamental difference from the original scheme is the fact that the block diagram has nested subproblems

\begin{equation}
\label{subTasks}
\phi_i(u_1,\dots,u_i)=\min_{u_{i+1}\in D_{i+1}}\phi_{i+1}(u_1,\dots,u_i,u_{i+1}),1\leqslant i\leqslant M-1
\end{equation}

which are multidimensional, and to solve them, a method of reducing dimensionality based on Peano curves can be applied.

@@

The solution to the resulting set of subproblems (\ref{subTasks}) can be organized in various ways.

The obvious method (elaborated in detail in \cite{Ref23},\cite{Ref24},\cite{Ref25} for the nested optimization scheme and in \cite{Ref26},\cite{Ref27} for the block nested optimization scheme) is based on solving subproblems in accordance with the recursive order of their generation. However, there is a significant loss of information about the target function.

Another approach is an adaptive scheme in which all subproblems are solved simultaneously, which makes it possible to more fully take into account information about a multidimensional problem and thereby to speed up the process of solving it.

In the case of one-dimensional subproblems, this approach was theoretically substantiated and tested in \cite{Ref28}, \cite{Ref29}, \cite{Ref30}, and in the article [] a generalization of the adaptive scheme for multidimensional subproblems was proposed.

   The adaptive dimensionality reduction scheme changes the order in which subproblems are solved: they will be solved not one by one (in accordance with their hierarchy in the task tree), but simultaneously, i.e., there will be a number of subproblems that are in the process of being solved. Under the new scheme:

•	to calculate the value of the function of the \(i\)-th level from (\ref{subTasks}), a new level \(i+1\) problem is generated in which tests are carried out, after which a new generated problem is included in the set of existing problems to be solved;

•	the iteration of the global search consists in choosing \(P\) (the most promising) problem from the set of existing problems in which tests are carried out; points for new tests are determined in accordance with the algorithm of subsection 1;

•	the minimum values of functions from (\ref{subTasks}) are their current estimates based on the accumulated search information. .

A brief description of the main steps of a block adaptive dimensionality reduction scheme is as follows.

Let the nested subproblems in the form (\ref{subTasks}) be solved using the global search algorithm described in subsection 1. Then each subproblem (\ref{subTasks}) can be assigned a numerical value called the characteristic of this problem. As such, we can take the maximum characteristic of the intervals formed in this problem – \(R(t)\). In accordance with the rule for calculating characteristics, the higher the value of the characteristic, the more promising the subproblem is in the continued search for the global minimum of the original problem (\ref{task}). Therefore, at each iteration, subproblems with the maximum characteristic are selected for conducting the next test. The test either leads to the calculation of the value of the objective function \(\phi_1(y)\) (if the selected subproblem belonged to the level \(j=M\)), or generates new subproblems according to (\ref{subTasks}) for \(j \geqslant  M-1\). In the latter case, the newly generated problems are added to the current set of problems, their characteristics are calculated, and the process is repeated. The optimization process is completed when the root problem satisfies the condition for stopping the algorithm that solves this problem. Some results pointing in this direction are presented in \cite{Ref31}, \cite{Ref32}.

@@

In relation to global optimization methods, an operation that can be effectively implemented on the GPU is the parallel calculation of many values of the objective function at once. Naturally, this requires implementing a procedure for calculating the value of a function on the GPU. Data transfers from the CPU to the GPU will be minimal: one need only transfer the coordinates of the test points to the GPU, and get back the function values at these points. Functions that determine the processing of test results in accordance with the algorithm, and require work with large amounts of accumulated search information, can be effectively implemented on the CPU.

   The general scheme for organizing calculations using the GPU is shown in \ref{fig:GPU_Algorithm}. In accordance with this scheme, steps 1-4 of the parallel global search algorithm are performed on the CPU. The coordinates of the P test points calculated in step 4 of the algorithm are accumulated in the intermediate buffer and then transmitted to the GPU. On the GPU, the function values are calculated at these points, after which the test results (again through the intermediate buffer) are transferred to the CPU.

   \begin{figure}[h!]
    \centering
		\includegraphics[width=1.0\textwidth]{GPU_Algorithm.eps}
		\caption{Scheme of information exchanges in the GPU algorithm}\label{fig:GPU_Algorithm}
\end{figure}

@@

To organize parallel calculations, we will use a set of evolvents and a block adaptive dimensionality reduction scheme. We take a small number of nesting levels, 2 to 3, in which the original large dimensionality problem is divided into 2-3 nested lower dimensional subproblems. We will use multiple sweeps only at the upper level of nesting, corresponding to the u1 variable block. This subproblem will be reduced to a set of one-dimensional problems that will be solved in parallel, each in a separate process. Test results at point x obtained for the problem solved by a specific processor are interpreted as the test results in the remaining problems (at the corresponding points \(u^1_1,… , u^s_1\) )
Then, applying an adaptive scheme to solve the nested subproblems (\ref{subTasks}), we get a parallel algorithm with a wide degree of variability.

\ref{fig:Parallel_Computing} shows the general scheme of organizing calculations using several cluster nodes and several GPUs. In accordance with this scheme, nested subproblems  ${{\phi }_{i}}({{u}_{1}},...,{{u}_{i}})=\underset{{{u}_{i+1}}\in {{D}_{i+1}}}{\mathop{\min }}\,{{\phi }_{i+1}}({{u}_{1}},...,{{u}_{i}},{{u}_{i+1}})$  with \(i=1,\ldots ,M-2\) are solved using only the CPU. The values of the optimized function are not calculated directly in these subproblems: the calculation the function value  ${{\phi }_{i}}({{u}_{1}},...,{{u}_{i}})$ is a solution to the minimization problem at the next level.

The subproblem of the last \((M - 1)\) level \({{\phi }_{M-1}}({{u}_{1}},...,{{u}_{M-1}})=\underset{{{u}_{M}}\in {{D}_{M}}}{\mathop{\min }}\,{{\phi }_{M}}({{u}_{1}},...,{{u}_{M}})\) differs from all the previous subproblems – it calculates the values of the optimized function, since ${{\phi }_{M}}({{u}_{1}},...,{{u}_{M}})=\phi ({{y}_{1}},...,{{y}_{N}})$. This subproblem transfers between the CPU and the GPU.

   \begin{figure}[h!]
    \centering
		\includegraphics[width=1.0\textwidth]{Parallel_Computing.eps}
		\caption{Diagram of parallel computing on a cluster}\label{fig:Parallel_Computing}
\end{figure}

@@

The numerical experiments were carried out using the Lomonosov supercomputer (Lomonosov Moscow State University). Each supercomputer node included two quad-core processors, Intel Xeon X5570 2.93GHz and 12 Gb RAM. When building the Globalizer system for running on the Lomonosov supercomputer, the GCC 4.3.0 compiler, CUDA 6.5 and Intel MPI 2017 were used.

Note that well-known test problems from the field of multidimensional global optimization are characterized by a short time span for calculating the values of the objective function. Therefore, in order to simulate the computational complexity inherent in applied optimization problems, the calculation of the objective function in all experiments was complicated by additional calculations that do not change the form of the function and the location of its minima (summing the segment of the Taylor series).

In the study \cite{Ref15}, \cite{Ref16}, a GKLS generator is described that allows one to generate multiextremal optimization problems with previously known properties: the number of local minima, the size of their regions of attraction, the global minimum point, the value of the function in it, etc.

Below are the results of a numerical comparison of three sequential algorithms -- DIRECT \cite{Ref15}, DIRECTl \cite{Ref17} and Global Search Algorithm (GSA); (results of the first two algorithms are given in the article \cite{Ref18}). A numerical comparison was carried out on the Simple and Hard function classes of dimension 4 and 5 from \cite{Ref18}. The Global minimum $y*$ was considered found if the algorithm generated a test point $y^k$ in the $\delta$-neighborhood of the global minimum, i.e. $\left\| {{y}^{k}}-{{y}^{*}} \right\|\le \delta $. Also, the size of the neighborhood was chosen (in accordance with \cite{Ref18}) as $\delta =\left\| b-a \right\|\sqrt[N]{\Delta }$, here $N$ is the dimension of the problem to be solved, a и b are the boundaries of the search domain $D$, the parameter  $\Delta $=10-6 for N=4 and $\Delta =10-7$ for $N=5$. When using the AGP method for the Simple class, the parameter $r=4.5$ was selected, for the Hard class - $r=5.6$; the parameter for constructing the Peano curve was fixed, $m=10$. The maximum number of iterations allowed was $Kmax = 1 000 000$.

\begin{table}
    \centering
    \begin{tabular}{|c|p{3cm}|c|c|c|}
    \hline
    \(N\) & Problem class & DIRECT & DIRECT\(l\) & GSA \\ \hline
    4 & Simple & >47282 (4) & 18983 & 11953\\ \cline{2-5}
      & Hard & >95708 (7) & 68754 & 25263\\ \hline
    5 & Simple & >16057 (1) & 16758 & 15920\\ \cline{2-5}
      & Hard & >217215 (16) & >269064 (4) & >148342 (4)\\ \hline
    \end{tabular}
    \caption{Average number of iterations}
    \label{table:average_iters}
\end{table}

   Table  \ref{table:average_iters} shows the average number of iterations, $kav$, that the method performed when solving a series of problems from these classes. The symbol ">" reflects a situation where not all problems of the class were solved by any method whatsoever. This means that the algorithm was halted because the maximum allowed number of $Kmax$ iterations was reached. In this case, the value $Kmax = 1 000 000$ was used to calculate the average value of the number of iterations, $kav$, which corresponds to the lower estimate of this average value. The number of unsolved problems is indicated in parentheses.

  As can be seen from Table \ref{table:average_iters}, the sequential GSA surpasses the DIRECT and DIRECTl methods in all classes of problems in terms of the average number of iterations. At the same time, in the 5-Hard class, none of the methods solved all the problems: DIRECT failed to solve 16 problems, DIRECTl and GSA -- 4 problems each.

\begin{table}
    \centering
    \begin{tabular}{|c|p{3cm}|c|c|c|}
    \hline
    \(N\) & Problem class & P=2 & P=4 & P=8 \\ \hline
    4 & Simple  & 3,49 & 8,22  &  \\ \cline{2-5}
      & Hard    & 2,35 & 7,65  &  \\ \hline
    5 & Simple  & 1,87 & 6,59  &  \\ \cline{2-5}
      & Hard    & 1,48 & 10,80  &   \\ \hline
    \end{tabular}
    \caption{Acceleration of iterations on the CPU}
    \label{table:average_iters_AS_CPU}
\end{table}

\begin{table}
    \centering
    \begin{tabular}{|c|p{3cm}|c|c|c|}
    \hline
    \(N\) & Problem class & P=2 & P=4 & P=8 \\ \hline
    4 & Simple  & 1,71 & 2,01  &  \\ \cline{2-5}
      & Hard    & 1,17 & 1,90  &  \\ \hline
    5 & Simple  & 0,88 & 1,48  &  \\ \cline{2-5}
      & Hard    & 0,70 & 2,39  &   \\ \hline
    \end{tabular}
    \caption{Time acceleration on the CPU}
    \label{table:average_Time_AS_CPU}
\end{table}

   Let us now evaluate the acceleration achieved using parallel GSA using an adaptive dimensionality reduction scheme based on the number $P$ of cores used. Table \ref{table:average_iters_AS_CPU} shows the acceleration by iterations, and Table \ref{table:average_Time_AS_CPU} shows the time acceleration obtained when solving a series of problems on the CPU, compared to GSA.

\begin{table}
    \centering
    \begin{tabular}{|c|p{3cm}|c|c|c|}
    \hline
    \(N\) & Problem class & P=2 & P=4 & P=8 \\ \hline
    4 & Simple &   6,62 & 21,46  &  \\ \cline{2-5}
      & Hard   &   3,23 & 9,97  &  \\ \hline
    5 & Simple &   3,37 & 11,20  &  \\ \cline{2-5}
      & Hard   &   2,52 & 8,54  &   \\ \hline
    \end{tabular}
    \caption{Time acceleration on the CPU}
    \label{table:average_iters_MMAS_CPU}
\end{table}

\begin{table}
    \centering
    \begin{tabular}{|c|p{3cm}|c|c|c|}
    \hline
    \(N\) & Problem class & P=2 & P=4 & P=8 \\ \hline
    4 & Simple &   3,22 & 5,23  &  \\ \cline{2-5}
      & Hard   &   1,48 & 2,42  &  \\ \hline
    5 & Simple &   1,30 & 1,97  &  \\ \cline{2-5}
      & Hard   &   0,68 & 0,88  &   \\ \hline
    \end{tabular}
    \caption{Acceleration of iterations on the CPU}
    \label{table:average_Time_MMAS_CPU}
\end{table}

   Table \ref{table:average_iters_MMAS_CPU} shows the acceleration in iterations, and Table \ref{table:average_Time_MMAS_CPU} shows the acceleration in time of the algorithm that combines multiple sweeps and an adaptive scheme for solving a series of problems on the CPU, compared to the sequential launch of the GSA. Two scans and, accordingly, two processes were used; each process used $P$ threads, and calculations were performed on a single cluster node.

\begin{table}
    \centering
    \begin{tabular}{|c|p{3cm}|c|c|c|}
    \hline
    \(N\) & Problem class & P=64 & P=256 & P=1024 \\ \hline
    4 & Simple &   5,06 & 19,91 & 15,62  \\ \cline{2-5}
      & Hard   &   3,93 & 14,95 &   \\ \hline
    5 & Simple &   4,16 & 16,12 & 67,22  \\ \cline{2-5}
      & Hard   &   3,36 & 10,02 & 42,04   \\ \hline
    \end{tabular}
    \caption{Acceleration of iterations on the GPU}
    \label{table:average_iters_AS_GPU}
\end{table}

\begin{table}
    \centering
    \begin{tabular}{|c|p{3cm}|c|c|c|}
    \hline
    \(N\) & Problem class & P=64 & P=256 & P=1024 \\ \hline
    4 & Simple &   2,20 & 8,28 & 5,21  \\ \cline{2-5}
      & Hard   &   1,92 & 6,91 &   \\ \hline
    5 & Simple &   1,58 & 4,18 & 8,23  \\ \cline{2-5}
      & Hard   &   1,22 & 2,39 & 3,74   \\ \hline
    \end{tabular}
    \caption{Time acceleration on the GPU}
    \label{table:average_Time_AS_GPU}
\end{table}

\begin{table}
    \centering
    \begin{tabular}{|c|p{3cm}|c|c|c|}
    \hline
    \(N\) & Problem class & P=64 & P=256 & P=1024 \\ \hline
    4 & Simple &   11,32 & 38,96 & 128,39  \\ \cline{2-5}
      & Hard   &   6,83  & 31,52 &   \\ \hline
    5 & Simple &   4,99  & 39,37 & 80,56  \\ \cline{2-5}
      & Hard   &   4,61  & 17,28 & 113,61   \\ \hline
    \end{tabular}
    \caption{Acceleration of iterations on the GPU}
    \label{table:average_iters_MMAS_GPU}
\end{table}

\begin{table}
    \centering
    \begin{tabular}{|c|p{3cm}|c|c|c|}
    \hline
    \(N\) & Problem class & P=64 & P=256 & P=1024 \\ \hline
    4 & Simple &   2,44 & 8,01 & 20,45  \\ \cline{2-5}
      & Hard   &   1,63 & 7,04 &   \\ \hline
    5 & Simple &   0,77 & 3,97 & 2,37  \\ \cline{2-5}
      & Hard   &   0,61 & 0,96 & 2,11   \\ \hline
    \end{tabular}
    \caption{Time acceleration on the GPU}
    \label{table:average_Time_MMAS_GPU}
\end{table}

Tables \ref{table:average_iters_AS_GPU} and  \ref{table:average_Time_AS_GPU} show the acceleration obtained when solving a series of problems on the GPU using an adaptive scheme compared to a similar launch on the CPU using 4 threads. Tables \ref{table:average_iters_MMAS_GPU} and \ref{table:average_Time_MMAS_GPU} show the acceleration of the algorithm combining multiple sweeps and an adaptive circuit when solving a series of problems on the GPU, compared to an adaptive scheme on the CPU using 4 threads. Two sweeps and, accordingly, two processes were used; each process used P threads, and calculations were performed on a single cluster node.

\begin{table}
    \centering
    \begin{tabular}{|p{4cm}|c|c|c|}
    \hline
      & P=2 & P=4 & P=8 \\ \hline
    Acceleration of iterations&   30,79 & 92,68 & 596,97  \\ \hline
    Time acceleration         &   1,86  & 1,49  & 2,45    \\ \hline

    \end{tabular}
    \caption{Acceleration of iterations and time on the GPU}
    \label{table:GKLS6_average_MMAS_GPU}
\end{table}

Следующий эксперимент будет проведен на 6-ти мерных задачах GKLS. В таблице \ref{table:GKLS6_average_MMAS_GPU} приведено ускорение по времени и по итерациям полученное при решении серии задач на GPU с использованием алгоритма объединяющего множественные развертки и адаптивную схему по сравнению с адаптивной схемой на CPU использующей 4 потока. Эксперимент проводился на трех узлах кластера и спользовалось до 6144 нитей.

@@

%
% ---- Bibliography ----
%
\begin{thebibliography}{6}

\bibitem {Ref1}	Hwu, W. GPU Computing Gems: Emerald Edition — San Francisco: Morgan Kaufmann, 2011. — 886 p.
\bibitem {Ref2}	A.M. Ferreiro, J.A. Garcia, J.G. Lopez-Salas, C. Vazquez. An efficient implementation of parallel simulated annealing algorithm in GPUs // Journal of global optimization. — 2013. — V. 57, No 3. — P. 863–890.
\bibitem {Ref3}	J.M. Garcia-Martinez, E.M. Garzon, P.M. Ortigosa. A GPU implementation of a hybrid evolutionary algorithm: GPuEGO // Journal of super-computing. — 2014. — V. 70, No 2. — P. 684–695.
\bibitem {Ref4}	Langdon, W.B. Graphics processing units and genetic programming: an overview // Soft Computing. — 2011. — V. 15, No 8. — P. 1657–1669.
\bibitem {Ref5}	Ю.Г. Евтушенко, В.У. Малкова, А.А. Станевичюс. Параллельный поиск глобального экстремума функций многих переменных // Ж. вычисл. матем. и матем. физ. — 2009. — Т. 49, No2. — С. 255–269.
\bibitem {Ref6}	J. He, A. Verstak, L.T. Watson, M. Sosonkina. Design and implementation of a massively parallel version of DIRECT // Computational optimization and applications. — 2008. — V. 40, No 2. — P. 217–245.
\bibitem {Ref7}	R. Paulavicius, J. Žilinskas, A. Grothey. Parallel branch and bound for global optimization with combination of Lipschitz bounds // Optimization methods and software. — 2011. — V. 26, No 3. — P. 487–498.
\bibitem {Ref8}	V. Kindratenko (ed.). Numerical computations with GPUs — New York: Springer, 2014. — 405 p.
\bibitem {Ref9}	Strongin, R.G. Global Optimization with Non-convex Constraints. Sequential and Parallel Algorithms / R.G. Strongin, Ya.D. Sergeyev – Kluwer Academic Publishers, 2000. – 704 p.
\bibitem {Ref10}	Sergeyev, Ya.D. Introduction to global optimization exploiting space-filling curves / Ya.D. Sergeyev, R.G. Strongin, D. Lera – Springer, 2013. – 125 p.
\bibitem {Ref11}	Стронгин, Р.Г. Параллельные вычисления в задачах глобальной оптимизации / Р.Г. Стронгин, В.П. Гергель, В.А. Гришагин, К.А. Баркалов – М.: Издательство Московского университета, 2013. – 280 с.
